---
title: "P8105 Homework 2"
output: github_document
date: "2023-10-02"
---

# Problem 1
Goal: merging the FiveThirtyEight `pols-month.csv`, `unemployment.csv`, and `snp.csv` data into a single data frame using year and month as keys across datasets.

## Loading Libraries
First, loading the `tidyverse` library for use throughout the homework.
```{r library}
library(tidyverse)
```

## Cleaning `pols-month` Data
Now, cleaning the data in `pols-month.csv`:

* Using separate() to break up the variable mon into integer variables year, month, and day
* Replacing month number with month name
* Creating a president variable taking values gop and dem
* Removing prez_dem and prez_gop
* Removing the day variable

```{r cleaning pols-month}

pols_month = 
  read_csv("Data/pols-month.csv") |> 
  janitor::clean_names() |> 
  separate(mon, into = c("year", "month", "day"), sep = "-") |> 
  mutate(
    year = as.integer(year),
    month = as.integer(month),
    day = as.integer(day),
    president = ifelse(prez_dem > 0 , "dem", "gop"),
    month = case_match(
      month,
      1 ~ "January",
      2 ~ "February",
      3 ~ "March",
      4 ~ "April",
      5 ~ "May",
      6 ~ "June",
      7 ~ "July",
      8 ~ "August",
      9 ~ "September",
      10 ~ "October",
      11 ~ "November",
      12 ~ "December")
    ) |> 
  select(-starts_with("prez"), -day)

```

## Cleaning `snp` Data
Second, cleaning the data in `snp.csv` using a similar process:

* Using separate() to break up the variable mon into integer variables year, month, and day
* Replacing month number with month name
* Creating a president variable taking values gop and dem
* Removing prez_dem and prez_gop
* Removing the day variable
* Arrange according to year and month, and with those as the leading columns

```{r}

```

## Cleaning `unemployment` Data
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

## Merging the Data Sets
Join the datasets by merging snp into pols, and merging unemployment into the result.

## Describing the Data Sets
Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).


